---
title: "lab 6"
author: "Kat Mackay"
date: "2023-02-16"
output: html_document
  html_document:
    code_folding: show
---

```{r setup, echo = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(janitor)
library(palmerpenguins)

### Packages for cluster analysis
library(NbClust)
#install.packages("NbClust")
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
#install.packages("factoextra")
#install.packages("dendextend")
#install.packages("ggdendro")
```

#Intro to cluster analysis - k-means, heirarchical 

## Part 1: K-means clustering 
```{r}
ggplot(penguins) + 
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm,
                 color = species, shape = sex), 
             size = 3, alpha = 0.7) + 
  scale_color_manual(values = c('orange', 'cyan4', 'darkmagenta'))

ggplot(penguins) + 
  geom_point(aes(x = flipper_length_mm, 
                 y = body_mass_g, 
                 color = species, 
                 shape = sex),
             size = 3, alpha = 0.7) + 
  scale_color_manual(values = c('orange', 'cyan4', 'darkmagenta'))
```

### Create a complete, scaled version of the data 
```{r}
penguins_complete <- penguins %>% 
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm) 

penguins_scale <- penguins_complete %>% 
  select(ends_with('_mm'), body_mass_g) %>% 
  scale()

summary(penguins_scale)
```

### Estimate number of clusters 
```{r}
number_est <- NbClust(penguins_scale, 
                      min.nc = 2, max.nc = 10, 
                      method = 'kmeans')

fviz_nbclust(penguins_scale, FUNcluster = kmeans, method = 'wss', k.max = 10)
```

### Run some k-means
```{r}
set.seed(123)

penguins_km <- kmeans(penguins_scale, 
                      centers = 3, 
                      iter.max = 10, 
                      nstart = 25)

#penguins_km$size
#penguins_km$cluster

penguins_cl <- penguins_complete %>% 
  mutate(cluster_no = factor(penguins_km$cluster))
```

```{r}
ggplot(penguins_cl) + 
  geom_point(aes(x = flipper_length_mm, 
                 y = body_mass_g, 
                 color = cluster_no,
                 shape = species))


ggplot(penguins_cl) + 
  geom_point(aes(x = bill_length_mm, 
                 y = bill_depth_mm,  
                 color = cluster_no,
                 shape = species)) + 
             scale_color_manual(values = c('lightblue', 'lightpink', 'lightgreen'))

penguins_cl %>% 
  select(species, cluster_no) %>% 
  table()
  
```

## Hierarchical clustering 

### Start with complete linkage
```{r}
### Create distance matrix 
peng_dist <- dist(penguins_scale, method = 'euclidean')

### Hierarchical clustering (complete linkage)
peng_hc_complete <- hclust(peng_dist, method = 'complete')
## also: single, average, ward.D

plot(peng_hc_complete, cex = 0.6, hang = -1)

### Cut into three clusters 
peng_cut_hc <- cutree(peng_hc_complete, 3)
table(peng_cut_hc, penguins_complete$species)
# BLR uses multiple variables but uses it to create a predictive model based on the data to determine what we would expect to see from each species 
# know that the group has certain characteristics but we aren't sure what the characteristics are 
```

## World Bank data: read in and simplify 
```{r}
wb_env <- read_csv(here::here('data/wb_env.csv'))

wb_ghg_20 <- wb_env %>% 
  slice_max(n = 20, ghg)

summary(wb_ghg_20)

wb_scaled <- wb_ghg_20 %>% 
  select(3:7) %>% 
  scale()

summary(wb_scaled)

rownames(wb_scaled) <- wb_ghg_20$name  # allows us to use a column we had before as the names of the rows 

```

# Find the Euclidian distances 

```{r}
euc_distance <- dist(wb_scaled, method = 'euclidian')
euc_distance
```

### Perform hierarchical clustering with complete linkage
```{r}
hc_complete <- hclust(euc_distance, method = 'complete')

plot(hc_complete, cex = .6, hang = -1)
```

### Perform heirarchical clustering by single linkage
```{r}
hc_single <- hclust(euc_distance, method = 'single')

plot(hc_single, cex = .6, hang = -1)
```

### Make a tanglegram 
```{r}
dend_complete <- as.dendrogram(hc_complete)
dend_single <- as.dendrogram(hc_single)

tanglegram(dend_complete, dend_single)
#compares the way they were clustered together 
#countries that are being grouped or clustered in the same way based on the complete vs single linkage

entanglement(dend_complete, dend_single) #lower is better

untangle(dend_complete, dend_single, method = 'step1side') %>% 
  entanglement() #matched them better going across -- easier to see the differences between the two 

untangle(dend_complete, dend_single, method = 'step1side') %>% 
  tanglegram(common_subtrees_color_branches = TRUE) 
```
### Let's make a dendrogram in ggplot! 
```{r}
ggdendrogram(hc_complete, rotate = TRUE) + 
  theme_minimal() + 
  labs(x = 'Country')
```

